State of the Practice for Medical Imaging Software Based on Open Source Repositories

### 2024-05-28 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
Some edits to the SOP paper, reorg to match required section titles

We present the state of the practice for Medical Imaging (MI) software based on
data available in open source repositories. We selected 29 medical imaging
projects from 48 candidates and assessed 9 software qualities (installability,
correctness/ verifiability, reliability, robustness, usability, maintainability,
reusability, understandability, and visibility/transparency) by answering 108
questions for each software project. Based on the quantitative data, we ranked
the MI software with the Analytic Hierarchy Process (AHP). The top four software
products are \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, and
\textit{OHIF Viewer}.  Our ranking is mostly consistent with the community's
ranking, with four of our top five projects also appearing in the top five of a
list ordered by stars-per-year.  We observed 88\% of the documentation artifacts
recommended by research software development guidelines.  However, the current
state of the practice deviates from the existing guidelines because of the
rarity of some recommended artifacts (like test plans, requirements
specification, code of conduct, code style guidelines, product roadmaps, and
Application Program Interface (API) documentation).

@@

Not only do we wish to gain insight into the state of the practice for MI
software, we also wish to understand the development of research software in
general. We wish to understand the impact of the often cited gap, or chasm,
between software engineering and research software \citep{Kelly2007,
Storer2017}. Although scientists spend a substantial proportion of their working
hours on software development \citep{Hannay2009, Prabhu2011}, many developers
learn software engineering skills by themselves or from their peers, instead of
from proper training \citep{Hannay2009}. \citet{Hannay2009} observe that many
scientists showed ignorance and indifference to standard software engineering
concepts. For instance, according to a survey by \citet{Prabhu2011}, more than
half of their 114 subjects did not use a proper debugger when coding.

@@

\subsection{Methodology Overview}

@@

\section{Background} \label{ch_background}

To measure the existing MI software we need the definitions of the software
qualities that we will be assessing (Section \ref{sec_software_quality}). In our
assessment we rank the software packages for each quality; therefore, this
section also provides the background on our ranking process --- the Analytic
Hierarchy Process (Section \ref{sec_AHP}).

\subsection{Software Quality Definitions} \label{sec_software_quality}

Quality is defined as a measure of the excellence or worth of an entity.  As is
common practice, we do not think of quality as a single measure, but rather as a
set of measures.  That is, quality is a collection of different qualities, often
called ``ilities.''  Below we list the 10 qualities of interest for this study.
The order of the qualities follows the order used in \citet{GhezziEtAl2003},
which puts related qualities (like correctness and reliability) together.
Moreover, the order is roughly the same as the order developers consider
qualities in practice.

\begin{itemize}
	\item \textbf{Installability} The effort required for the installation
    and/or uninstallation of software in a specified environment
    \citep{ISO/IEC25010, lenhard2013measuring}.

	\item \textbf{Correctness \& Verifiability} A program is correct if it
    matches its specification \citep[p.\ 17]{GhezziEtAl2003}.  The specification
    can either be explicitly or implicitly stated.  The related quality of
    verifiability is the ease with which the software components or the
    integrated product can be checked to demonstrate its correctness.

	\item \textbf{Reliability} The probability of failure-free operation of a
	computer program in a specified environment for a specified time
	\citep{musa1987software}, \citep[p.\ 357]{GhezziEtAl2003}.

	\item \textbf{Robustness} Software possesses the characteristic of
	robustness if it behaves ``reasonably'' in two situations: i) when it
	encounters circumstances not anticipated in the requirements specification,
	and ii) when users violate the assumptions in its requirements specification 
	\citep[p.\ 19]{GhezziEtAl2003}, \citep{boehm2007software}.

	\item \textbf{Usability} ``The extent to which a product can be used by
	specified users to achieve specified goals with effectiveness, efficiency,
	and satisfaction in a specified context of use'' \citep{ISO/TR16982:2002,
	ISO9241-11:2018}.

	\item \textbf{Maintainability} The effort with which a software system or
	component can be modified to i) correct faults; ii) improve performance or
	other attributes; iii) satisfy new requirements
	\citep{IEEEStdGlossarySET1990, boehm2007software}.

	\item \textbf{Reusability} ``The extent to which a software component can be
	used with or without adaptation in a problem solution other than the one for
	which it was originally developed'' \citep{kalagiakos2003non}.

	\item \textbf{Understandability} ``The capability of the software product to
	enable the user to understand whether the software is suitable, and how it
	can be used for particular tasks and conditions of use'' \citep{iso2001iec}.

	\item \textbf{Visibility/Transparency} The extent to which all the steps
	of a software development process and the current status of it are conveyed
	clearly \citep[p.\ 32]{GhezziEtAl2003}.

\subsection{Analytic Hierarchy Process (AHP)} \label{sec_AHP}

@@

\section{Methodology} \label{ch_methods}

\begin{enumerate}
\item Identify list of representative software packages
(Section~\ref{sec_software_selection});
\item Measure (or grade) the selected software
(Section~\ref{sec_grading_software});
\item Answer the research questions (as given in Section~\ref{sec_motivation}).
\end{enumerate}

In the sections below we provide additional detail on the above steps, while
concurrently giving examples of how we applied the methodology to the MI domain.

\subsection{Interaction With Domain Expert} \label{sec_vet_software_list}

\subsection{List of Representative Software} \label{sec_software_selection}

We have a two-step process for selecting software packages: i) identify software
candidates in the chosen domain; and, ii) filter the list to remove less
relevant members \citep{SmithEtAl2021}.

@@

\subsection{Grading Software} \label{sec_grading_software}

For measuring each quality, we ask several questions and the typical answers are
among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
string, a date, a set of strings, etc. The grader assigns each quality an
overall score, between 1 and 10, based on all the previous questions.  Several
of the qualities use the word ``surface''.  This is to highlight that, for these
qualities in particular, the best that we can do is a shallow measure.  For
instance, we are not currently doing any experiments to measure usability.
Instead, we are looking for an indication that the developers considered
usability.  We do this by looking for cues in the documentation, like a getting
started manual, a user manual and a statement of expected user characteristics.
Below is a summary of how we assess adoption of best practices by measuring each
quality.

\item \textbf{Installability} We assess the following: 
\begin{inparaenum}[i)]
    \item existence and quality of installation instructions;
    \item the quality of the user experience via the ease of following
    instructions, number of steps, automation tools; and,
    \item whether there is a means to verify the installation.
\end{inparaenum}
If any problem interrupts the process of installation or uninstallation, we give
a lower score. We also record the Operating System (OS) used for the
installation test.

% move the definitions here; it seems repetitive the way it is now - maybe
% remove the definitions?
\item \textbf{Correctness \& Verifiability} We check each project to identify
any techniques used to ensure this quality, such as literate programming,
automated testing, symbolic execution, model checking, unit tests, etc. We also
examine whether the projects use Continuous Integration and Continuous Delivery
(CI/CD). For verifiability, we go through the documents of the projects to check
for the presence of requirements specifications, theory manuals, and getting
started tutorials. If a getting started tutorial exists and provides expected
results, we follow it to check the correctness of the output.

\item \textbf{Surface Reliability} We check the following: 
\begin{inparaenum}[i)]
    \item whether the software breaks during installation;
    \item the operation of the software following the getting started tutorial
    (if present);
    \item whether the error messages are descriptive; and,
    \item whether we can recover the process after an error.
\end{inparaenum}

\item \textbf{Surface Robustness} We check how the software handles
unexpected/unanticipated input. For example, we prepare broken image files for
MI software packages that load image files. We use a text file (.txt) with a
modified extension name (.dcm) as an unexpected/unanticipated input. We load a
few correct input files to ensure the function is working correctly before
testing the unexpected/unanticipated ones.

\item \textbf{Surface Usability} We examine the project's documentation,
checking for the presence of getting started tutorials and/or a user manual. We
also check whether users have channels to request support, such as an e-mail
address, or issue tracker. Our impressions of usability are based on our
interaction with the software during testing.  In general, an easy-to-use
graphical user interface will score high.

\item \textbf{Maintainability} We believe that the artifacts of a project,
including source code, documents, and building scripts, significantly influence
its maintainability. Thus, we check each project for the presence of such
artifacts as API documentation, bug tracker information, release notes, test
cases, and build scripts. We also check for the use of tools supporting issue
tracking and version control, the percentages of closed issues, and the
proportion of comment lines in the code.

\item \textbf{Reusability} We count the total number of code files for each
project. Projects with numerous components potentially provide more choices for
reuse. Furthermore, well-modularized code, which tends to have smaller parts in
separate files, is typically easier to reuse. Thus, we assume that projects with
more code files and fewer Lines of Code (LOC) per file are more reusable. We also
consider projects with API documentation as delivering better reusability.

\item \textbf{Surface Understandability} Given that time is a constraint, we
cannot look at all code files for each project; therefore, we randomly examine
10 code files for their understandability. We check the code's style within each
file, such as whether the identifiers, parameters, indentation, and formatting
are consistent, whether the constants (other than 0 and 1) are not hardcoded, and
whether the code is modularized. We also check the descriptive information for
the code, such as documents mentioning the coding standard, the comments in the
code, and the descriptions or links for details on algorithms in the code.

\item \textbf{Visibility/Transparency} To measure this quality, we check the
existing documents to find whether the software development process and
current status of a project are visible and transparent. We examine the
development process, current status, development environment, and release notes
for each project.
\end{itemize}

\section{Measurement Results} \label{ch_results}

@@

\section{Comparison to Community Ranking} \label{Sec_VsCommunityRanking}

@@

\section{Comparison Between MI and Research Software for Artifacts}
\label{Sec_CompareArtifacts}

@@

\section{Threats to Validity} \label{sec_threats_to_validity}

@@

\subsection{Reliability}

@@

\subsection{Construct Validity}

@@

\subsection{Internal Validity} \label{Sec_InternalValidity}

@@

\subsection{External Validity}

>>>>

We present the state of the practice for Medical Imaging (MI) software based on
data available in open source repositories. We selected 29 medical imaging
projects from 48 candidates and assessed 9 software qualities (installability,
correctness/ verifiability, reliability, robustness, usability, maintainability,
reusability, understandability, and visibility/transparency) by answering 108
questions for each software project. Based on the quantitative data, we ranked
the MI software with the Analytic Hierarchy Process (AHP). The top four software
products are \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, and
\textit{OHIF Viewer}.  Our ranking is consistent with the community's
ranking, with four of our top five projects also appearing in the top five of a
list ordered by stars-per-year.  We observed 88\% of the documentation artifacts
recommended by research software development guidelines.  However, the current
state of the practice deviates from the existing guidelines because of the
rarity of some recommended artifacts (like test plans, requirements
specification, code of conduct, code style guidelines, product roadmaps, and
Application Program Interface (API) documentation).

@@

Not only do we wish to gain insight into the state of the practice for MI
software, we also wish to understand the impact of the often cited gap, or chasm,
between software engineering and research software \citep{Kelly2007,
Storer2017}. Although scientists spend a substantial proportion of their working
hours on software development \citep{Hannay2009, Prabhu2011}, many developers
learn software engineering skills by themselves or from their peers, instead of
from proper training \citep{Hannay2009}. \citet{Hannay2009} observe that many
scientists showed ignorance and indifference to standard software engineering
concepts. For instance, according to a survey by \citet{Prabhu2011}, more than
half of their 114 subjects did not use a proper debugger when coding.

@@

\subsection{Methodology}

@@

\begin{enumerate}
\item Identify list of representative software packages
(Section~\ref{sec_software_selection});
\item Measure (or grade) the selected software
(Section~\ref{sec_grading_software}) on qualities using AHP;
\end{enumerate}

define software qualities and ahp.

\subsubsection{Interaction With Domain Expert} \label{sec_vet_software_list}

\subsubsection{Grading Software} \label{sec_grading_software}

For measuring each quality, we ask several questions and the typical answers are
among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
string, a date, a set of strings, etc. The grader assigns each quality an
overall score, between 1 and 10, based on all the previous questions.  Several
of the qualities use the word ``surface''.  This is to highlight that, for these
qualities in particular, the best that we can do is a shallow measure.  For
instance, we are not currently doing any experiments to measure usability.
Instead, we are looking for an indication that the developers considered
usability.  We do this by looking for cues in the documentation, like a getting
started manual, a user manual and a statement of expected user characteristics.
Details on how we measure each quality are provided in \citet{SmithEtAl2024}.

\subsubsection{Software Quality Definitions} \label{sec_software_quality}

Quality is defined as a measure of the excellence or worth of an entity.  As is
common practice, we do not think of quality as a single measure, but rather as a
set of measures.  That is, quality is a collection of different qualities, often
called ``ilities.''  For this study we selected 9 qualities to measure: .....
With the exception of installability, all the qualities are defined in
\citet{GhezziEtAl2003}. Installability is defined as the effort required for the
installation and/or uninstallation of software in a specified environment
\citep{ISO/IEC25010, lenhard2013measuring}.  Further details on the software
qualities can be found in \citet{SmithEtAl2024}.

\subsubsection{Analytic Hierarchy Process (AHP)} \label{sec_AHP}

@@

\section{Review} \label{ch_results}

In this section we review the selected software packages.  We used a two-step
process for selecting the software packages: i) identify software candidates in
the chosen domain; and, ii) filter the list to remove less relevant members
\citep{SmithEtAl2021}.

@@

@@

\section{Discussion}

The review section shows us how the ranking between the existing open source MI
software, but does our ranking match how practitioners in the field rank the
software.  how does our ranking compare to the software compare to other
compares compare to community ranking and look at artifacts.

\subsection{Comparison to Community Ranking} \label{Sec_VsCommunityRanking}

@@

\subsection{Comparison Between MI and Research Software for Artifacts}
\label{Sec_CompareArtifacts}

@@

\subsection{Threats to Validity} \label{sec_threats_to_validity}

@@

\subsubsection{Reliability}

@@

\subsubsection{Construct Validity}

@@

\subsubsection{Internal Validity} \label{Sec_InternalValidity}

@@

\subsubsection{External Validity}
